---
title: "Individual Auxillary Variable (IAV) Models"
date: "16 July 2025"
author: "Brian Baird and Smitom Borah"
output: html_notebook
---

# Description
This R notebook contains code necessary to run the individual auxiliary variable (IAV) models with and without binning coefficients. Coefficients are binned based on nutrient enrichment index, depth, and temperature.

# Packages
```{r echo=TRUE}
# List of required packages
packages <- c("nimble", "coda", "ggplot2", "tidyverse", "ggside")

# Install any that are not already installed
install_if_missing <- function(p) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p)
  }
}

invisible(lapply(packages, install_if_missing))

# Load all packages
lapply(packages, library, character.only = TRUE)
```
# Data
Let us load the data to run the model. The data has been post-processed from the data sources described in the associated research article.
```{r}
# load data
bnla_final <- read.csv("data/bnla_final.csv")
```

# Preliminary models based on limiting nutrient
## Binning based on enrichment index
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.cr ~dnorm(20, sd = 5)
  sigma ~ dunif(0, 10)
  s.reint ~ dunif(0, 10)
  s.rebR ~ dunif(0,10)
  s.reln ~ dunif(0,10)
  s.lake ~ dunif(0,10)
  
  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i],x.tn[i]/(beta_bR[s.j[i]]))
    y[i] ~ dnorm(beta_ln[s.j[i]]*log(x.ln[i]) + beta_int[s.j[i]] + relake[s.k[i]], sd = sigma) 
  }
  for(j in 1:m) {
    beta_bR[j] <- b.cr + rebR[j]
    beta_int[j] <- b.0 + reint[j]
    beta_ln[j] <- b.tp + reln[j]
    reint[j] ~ dnorm(0, sd = s.reint)
    rebR[j] ~ dnorm(0, sd = s.rebR)
    reln[j] ~ dnorm(0, sd = s.reln)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
m=length(unique(bnla_final$eutro_bin)); m
p=length(unique(bnla_final$specific_lake_bin)); p

# Prepare bnla_finala for Nimble model
constants <- list(n = n,
                  m = m,
                  p = p,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  s.j = bnla_final$eutro_bin,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.cr = 15, sigma = 1, s.reint = 1, s.rebR = 1, s.reln = 1, s.lake = 1, relake=rep(0,p), reint=rep(0,m), rebR=rep(0,m), reln=rep(0,m))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("beta_bR", "beta_int", "beta_ln", "relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_nei <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_nei

# Summary of Output
summary(samples$samples) 
summary(samples$samples2[,1:15]) 

# Rhat Calculations
gelman.diag(samples$samples) 
gelman.diag(samples$samples2[,1:15]) 

# Make sample statistics a df mean values can be called directly
samp2 <- data.frame(summary(samples$samples2[,1:15])$statistics)

# Store lists of parameter coefficients
beta_bR_means <- samp2$Mean[1:5]; beta_bR_means
beta_int_means <- samp2$Mean[6:10]; beta_int_means
beta_tp_means <- samp2$Mean[11:15]; beta_tp_means

# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/(beta_bR_means[constants$s.j])) #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = beta_int_means[constants$s.j] + log(x.ln)*(beta_tp_means[constants$s.j])

# Calculate model performance
## R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

## RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```


## Binning based on depth

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.cr ~dnorm(20, sd = 5)
  sigma ~ dunif(0, 10)
  s.reint ~ dunif(0, 10)
  s.rebR ~ dunif(0,10)
  s.reln ~ dunif(0,10)
  s.lake ~ dunif(0,10)
  
  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i],x.tn[i]/(beta_bR[s.j[i]]))
    y[i] ~ dnorm(beta_ln[s.j[i]]*log(x.ln[i]) + beta_int[s.j[i]] + relake[s.k[i]], sd = sigma) 
  }
  for(j in 1:m) {
    beta_bR[j] <- b.cr + rebR[j]
    beta_int[j] <- b.0 + reint[j]
    beta_ln[j] <- b.tp + reln[j]
    reint[j] ~ dnorm(0, sd = s.reint)
    rebR[j] ~ dnorm(0, sd = s.rebR)
    reln[j] ~ dnorm(0, sd = s.reln)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
m=length(unique(bnla_final$depth_bin)); m
p=length(unique(bnla_final$specific_lake_bin)); p

# Prepare data for Nimble model
constants <- list(n = n, 
                  m = m, 
                  p = p,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  s.j = bnla_final$depth_bin,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.cr = 15, sigma = 1, s.reint = 1, s.rebR = 1, s.reln = 1, s.lake = 1, relake=rep(0,p), reint=rep(0,m), rebR=rep(0,m), reln=rep(0,m))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("beta_bR", "beta_int", "beta_ln", "relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_depth <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_depth

# Summary of Output
summary(samples$samples) 
summary(samples$samples2[,1:15]) 

# Rhat Calculations
gelman.diag(samples$samples) 
gelman.diag(samples$samples2[,1:15]) 

# Make sample statistics a df mean values can be called directly
samp2 <- data.frame(summary(samples$samples2[,1:15])$statistics)

# Store lists of parameter coefficients
beta_bR_means <- samp2$Mean[1:5]; beta_bR_means
beta_int_means <- samp2$Mean[6:10]; beta_int_means
beta_tp_means <- samp2$Mean[11:15]; beta_tp_means

# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/(beta_bR_means[constants$s.j])) #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = beta_int_means[constants$s.j] + log(x.ln)*(beta_tp_means[constants$s.j])

# Calculate model performance
## R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

## RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```


## Binning based on temperature
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.cr ~dnorm(20, sd = 5)
  sigma ~ dunif(0, 10)
  s.reint ~ dunif(0, 10)
  s.rebR ~ dunif(0,10)
  s.reln ~ dunif(0,10)
  s.lake ~ dunif(0,10)
  
  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i],x.tn[i]/(beta_bR[s.j[i]]))
    y[i] ~ dnorm(beta_ln[s.j[i]]*log(x.ln[i]) + beta_int[s.j[i]] + relake[s.k[i]], sd = sigma) 
  }
  for(j in 1:m) {
    beta_bR[j] <- b.cr + rebR[j]
    beta_int[j] <- b.0 + reint[j]
    beta_ln[j] <- b.tp + reln[j]
    reint[j] ~ dnorm(0, sd = s.reint)
    rebR[j] ~ dnorm(0, sd = s.rebR)
    reln[j] ~ dnorm(0, sd = s.reln)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
m=length(unique(bnla_final$temp_bin)); m
p=length(unique(bnla_final$specific_lake_bin)); p

# Prepare data for Nimble model
constants <- list(n = n, 
                  m = m, 
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  s.j = bnla_final$temp_bin,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.cr = 15, sigma = 1, s.reint = 1, s.rebR = 1, s.reln = 1, s.lake = 1, relake=rep(0,p), reint=rep(0,m), rebR=rep(0,m), reln=rep(0,m))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("beta_bR", "beta_int", "beta_ln", "relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_temp <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_temp

# Summary of Output
summary(samples$samples) 
summary(samples$samples2[,1:15]) 


# Rhat Calculations
gelman.diag(samples$samples) 
gelman.diag(samples$samples2[,1:15]) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)
samp2 <- data.frame(summary(samples$samples2[,1:15])$statistics)

# Store lists of parameter coefficients
beta_bR_means <- samp2$Mean[1:5]
beta_int_means <- samp2$Mean[6:10]
beta_tp_means <- samp2$Mean[11:15]


# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/(beta_bR_means[constants$s.j])) #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = beta_int_means[constants$s.j] + log(x.ln)*(beta_tp_means[constants$s.j]) 

# Calculate R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

# Calculate RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```

# Primilinary models based on dual nutrients
## Binning based on enrichment index
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.tn ~dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.reint ~ dunif(0, 10)
  s.retn ~ dunif(0,10)
  s.retp ~ dunif(0,10)
  s.lake ~ dunif(0,10)
  
  for(i in 1:n) {
    y[i] ~ dnorm((beta_tp[s.j[i]])*log(x.tp[i]) + (beta_tn[s.j[i]])*log(x.tn[i]) + beta_int[s.j[i]] + relake[s.k[i]], sd = sigma)
  }
  for(j in 1:m) {
    beta_tn[j] <- b.tn + retn[j]
    beta_int[j] <- b.0 + reint[j]
    beta_tp[j] <- b.tp + retp[j]
    reint[j] ~ dnorm(0, sd = s.reint)
    retn[j] ~ dnorm(0, sd = s.retn)
    retp[j] ~ dnorm(0, sd = s.retp)
  }
  for(k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
m=length(unique(bnla_final$eutro_bin)); m
p=length(unique(bnla_final$specific_lake_bin)); p

# Prepare data for Nimble model
constants <- list(n = n, 
                  m = m,
                  p = p,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  s.j = bnla_final$eutro_bin,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.tn = 1, sigma = 1, s.reint = 1, s.retn = 1, s.retp = 1, s.lake = 1, reint=rep(0,m), retn=rep(0,m), retp=rep(0,m), relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("beta_tn", "beta_int", "beta_tp", "relake"), enableWAIC = TRUE)
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_dual.nei <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_dual.nei

# Summary of Output
summary(samples$samples) 
summary(samples$samples2[,1:15]) 

# Rhat Calculations
gelman.diag(samples$samples) 
gelman.diag(samples$samples2[,1:15]) 

### Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)
samp2 <- data.frame(summary(samples$samples2[,1:15])$statistics)

# Store lists of parameter coefficients
beta_tn_means <- samp2$Mean[6:10]
beta_int_means <- samp2$Mean[1:5]
beta_tp_means <- samp2$Mean[11:15]

# Calculate Predictions
y.hat = (beta_tp_means[constants$s.j])*log(constants$x.tp) + (beta_tn_means[constants$s.j])*log(constants$x.tn) + beta_int_means[constants$s.j] 

# Calculate R2 and WAIC
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2

# Calculate RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```

## Binning based on depth
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.tn ~dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.reint ~ dunif(0, 10)
  s.retn ~ dunif(0,10)
  s.retp ~ dunif(0,10)
  s.lake ~ dunif(0,10)
  
  for(i in 1:n) {
    y[i] ~ dnorm((beta_tp[s.j[i]])*log(x.tp[i]) + (beta_tn[s.j[i]])*log(x.tn[i]) + beta_int[s.j[i]] + relake[s.k[i]], sd = sigma)
  }
  for(j in 1:m) {
    beta_tn[j] <- b.tn + retn[j]
    beta_int[j] <- b.0 + reint[j]
    beta_tp[j] <- b.tp + retp[j]
    reint[j] ~ dnorm(0, sd = s.reint)
    retn[j] ~ dnorm(0, sd = s.retn)
    retp[j] ~ dnorm(0, sd = s.retp)
  }
  for(k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
m=length(unique(bnla_final$depth_bin)); m
p=length(unique(bnla_final$specific_lake_bin)); p

# Prepare data for Nimble model
constants <- list(n = n, 
                  m = m,
                  p = p,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  s.j = bnla_final$depth_bin,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.tn = 1, sigma = 1, s.reint = 1, s.retn = 1, s.retp = 1, s.lake = 1, reint=rep(0,m), retn=rep(0,m), retp=rep(0,m), relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("beta_tn", "beta_int", "beta_tp", "relake"), enableWAIC = TRUE)
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_dual.depth <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_dual.depth

# Summary of Output
summary(samples$samples) 
summary(samples$samples2[,1:15]) 

# Rhat Calculations
gelman.diag(samples$samples) 
gelman.diag(samples$samples2[,1:15]) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)
samp2 <- data.frame(summary(samples$samples2[,1:15])$statistics)

# Store lists of parameter coefficients
beta_tn_means <- samp2$Mean[6:10]
beta_int_means <- samp2$Mean[1:5]
beta_tp_means <- samp2$Mean[11:15]


# Calculate Predictions
y.hat = (beta_tp_means[constants$s.j])*log(constants$x.tp) + (beta_tn_means[constants$s.j])*log(constants$x.tn) + beta_int_means[constants$s.j] 

# Calculate R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

# Calculate RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```

## Binning based on temperature

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.tn ~dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.reint ~ dunif(0, 10)
  s.retn ~ dunif(0,10)
  s.retp ~ dunif(0,10)
  s.lake ~ dunif(0,10)
  
  for(i in 1:n) {
    y[i] ~ dnorm((beta_tp[s.j[i]])*log(x.tp[i]) + (beta_tn[s.j[i]])*log(x.tn[i]) + beta_int[s.j[i]] + relake[s.k[i]], sd = sigma)
  }
  for(j in 1:m) {
    beta_tn[j] <- b.tn + retn[j]
    beta_int[j] <- b.0 + reint[j]
    beta_tp[j] <- b.tp + retp[j]
    reint[j] ~ dnorm(0, sd = s.reint)
    retn[j] ~ dnorm(0, sd = s.retn)
    retp[j] ~ dnorm(0, sd = s.retp)
  }
  for(k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
m=length(unique(bnla_final$temp_bin)); m
p=length(unique(bnla_final$specific_lake_bin)); p

# Prepare data for Nimble model
constants <- list(n = n, 
                  m = m,
                  p = p,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  s.j = bnla_final$temp_bin,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.tn = 1, sigma = 1, s.reint = 1, s.retn = 1, s.retp = 1, s.lake = 1, reint=rep(0,m), retn=rep(0,m), retp=rep(0,m), relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("beta_tn", "beta_int", "beta_tp", "relake"), enableWAIC = TRUE) # REMEMBER TO SET MONITORS
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_dual.temp <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_dual.temp

# Summary of Output
summary(samples$samples) 
summary(samples$samples2[,1:15]) 

# Rhat Calculations
gelman.diag(samples$samples) 
gelman.diag(samples$samples2[,1:15]) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)
samp2 <- data.frame(summary(samples$samples2[,1:15])$statistics)

# Store lists of parameter coefficients
beta_tn_means <- samp2$Mean[6:10]
beta_int_means <- samp2$Mean[1:5]
beta_tp_means <- samp2$Mean[11:15]

# Calculate Predictions
y.hat = (beta_tp_means[constants$s.j])*log(constants$x.tp) + (beta_tn_means[constants$s.j])*log(constants$x.tn) + beta_int_means[constants$s.j] 

# Calculate R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

# Calculate RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```


# Preliminary model with no bins
## Based on limiting nutrient
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.cr ~dnorm(20, sd = 5)
  sigma ~ dunif(0, 10)
  
  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i],x.tn[i]/(b.cr))
    y[i] ~ dnorm(b.0 + b.tp*log(x.ln[i]), sd = sigma) 
  }

})

# Dimension Assignment
n=length(bnla_final$chl); n

# Prepare data for Nimble model
constants <- list(n = n, 
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.cr = 15, sigma = 1)

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, enableWAIC = TRUE)
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_LimNut.noBins <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_LimNut.noBins

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)

# Store lists of parameter coefficients          
bR_means <- samp1$Mean[2]
int_means <- samp1$Mean[1]
tp_means <- samp1$Mean[3]

# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/(bR_means)) #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = int_means + log(x.ln)*(tp_means)

# Calculate R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

# Calculate RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```



## Based on dual nutrient
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model Formulation
code <- nimbleCode({
  b.0 ~ dnorm(0, sd = 10)
  b.tp ~ dnorm(0, sd = 10)
  b.tn ~dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)

  
  for(i in 1:n) {
    y[i] ~ dnorm((b.tp)*log(x.tp[i]) + (b.tn)*log(x.tn[i]) + b.0, sd = sigma)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n

# Prepare data for Nimble model
constants <- list(n = n, 
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn)
yt = log(bnla_final$chl) # Observed ln(chl) values
data <- list(y = yt)
inits <- list(b.0 = 1, b.tp = 1, b.tn = 1, sigma = 1)

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 50000 # Desired number of iterations
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 10000) 

# Save model
samples_DualNut.noBins <- samples
```

Here is the performance of the model:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Model
samples <- samples_DualNut.noBins

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)

# Store lists of parameter coefficients
tn_means <- samp1$Mean[2]
int_means <- samp1$Mean[1]
tp_means <- samp1$Mean[3]

# Calculate Predictions
y.hat = (tp_means)*log(constants$x.tp) + (tn_means)*log(constants$x.tn) + int_means

# Calculate R2
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 

# Calculate RMSE
rmse <- sqrt(mean(y.er^2)); rmse
```














