---
title: "Backward selection of variables for the MAV model"
date: "23 July 2025"
author: "Brian Baird and Smitom Borah"
output: html_notebook
---

# Description
This notebook contain coded used during the backwards variable selection (BVS) process for the MAV model. First, cleaned NLA data is loaded. Then, code is provided for a model including all significant candidate variables identified in the IAV models. Sequential models are created where insignificant parameters were removed one-by-one until a final model where all parameters are significant is achieved. 

# Packages
```{r echo=TRUE}
# List of required packages
packages <- c("nimble", "coda", "ggplot2", "tidyverse", "ggside", "psych")

# Install any that are not already installed
install_if_missing <- function(p) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p)
  }
}

invisible(lapply(packages, install_if_missing))

# Load all packages
lapply(packages, library, character.only = TRUE)
```

# Data
Let us load the data to run the model. The data has been post-processed from the data sources described in the associated research article.
```{r}
# load data
bnla_final <- read.csv("data/bnla_final.csv")

# Add observed TN:TP to dataset
bnla_final <- mutate(bnla_final, tn_tp = tn / tp)
lake_mean_n_p <- bnla_final %>%
  group_by(specific_lake_bin) %>%
  summarize(mean_lake_tn_tp = mean(tn_tp)) %>%
  ungroup()
```

# Backwards Variable Selection of MAV model
Initial model with all candidate variables
```{r Initial model with all candidate variables}
# Model Formulation
code <- nimbleCode({
  b0.int ~ dnorm(0, sd = 10)
  b0.int.depth ~ dnorm(0, sd = 10)
  b0.R ~ dnorm(20, sd = 10)
  b1.R.depth ~ dnorm(0, sd = 10)
  b1.R.tp ~ dnorm(0, sd = 10)
  b1.R.temp ~ dnorm(0, sd = 10)
  b0.slope ~ dnorm(0, sd = 10)
  b1.slope.temp ~ dnorm(0, sd = 10)
  b1.slope.depth ~ dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.lake ~ dunif(0,10)

  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i] , x.tn[i]/max((b1.R.depth * (log(x.depth[i]) - mu.log.depth) + b1.R.tp * (x.eutro[i]) + b0.R + b1.R.temp * (x.T[i] - mu.T)), 1))   ### Using original scale for slope of temperature (slope coefficient)
    y[i] ~ dnorm((b1.slope.temp * (x.T[i] - mu.T) + b1.slope.depth * (log(x.depth[i]) - mu.log.depth) + b0.slope)*log(x.ln[i]) + b0.int.depth * (log(x.depth[i]) - mu.log.depth) + b0.int + relake[s.k[i]], sd = sigma)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
p=length(unique(bnla_final$specific_lake_bin)); p
mu.T = mean(bnla_final$avg_temp)
mu.log.depth = mean(log(bnla_final$INDEX_SITE_DEPTH))
mu.log.tp = mean(log(bnla_final$tp))
mu.log.tp.2 = mean(log(bnla_final$tp)^2)

# Prepare data for Nimble model
constants <- list(n = n, 
                  p = p,
                  mu.T = mu.T,
                  mu.log.depth = mu.log.depth,
                  mu.log.tp = mu.log.tp,
                  mu.log.tp.2 = mu.log.tp.2,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  x.T = bnla_final$avg_temp,
                  x.eutro = bnla_final$log_eutro,
                  x.depth = bnla_final$INDEX_SITE_DEPTH,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl)
data <- list(y = yt)
inits <- list(b0.int = 1, b0.int.depth = 1, b1.R.temp = 1, b1.R.depth = 1, b1.R.tp = 1, b0.R = 1, b1.slope.temp = 1, b1.slope.depth = 1, b0.slope = 1, sigma = 1, s.lake = 1, relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 60000 
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 20000) 

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)

# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/pmax((samp1["b1.R.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) + samp1["b1.R.tp","Mean"] * (constants$x.eutro) +  samp1["b1.R.temp","Mean"] * (constants$x.T - mu.T) + samp1["b0.R", "Mean"]), 1))  #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = (samp1["b0.int", "Mean"] + samp1["b0.int.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) +  log(x.ln)*(samp1["b1.slope.temp","Mean"] * (constants$x.T - mu.T) + samp1["b1.slope.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) +  samp1["b0.slope","Mean"])) 

# Calculate R2 rmse
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 
rmse <- sqrt(mean(y.er^2)); rmse
```



## BVS 1
```{r BVS #1}
# Model Formulation
code <- nimbleCode({
  b0.int ~ dnorm(0, sd = 10)
  b0.int.depth ~ dnorm(0, sd = 10)
  b0.R ~ dnorm(20, sd = 10)
  b1.R.depth ~ dnorm(0, sd = 10)
  b1.R.tp ~ dnorm(0, sd = 10)
  b0.slope ~ dnorm(0, sd = 10)
  b1.slope.temp ~ dnorm(0, sd = 10)
  b1.slope.depth ~ dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.lake ~ dunif(0,10)

  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i] , x.tn[i]/max((b1.R.depth * (log(x.depth[i]) - mu.log.depth) + b1.R.tp * (x.eutro[i]) + b0.R), 1))  ### Using original scale for slope of temperature (slope coefficient)
    y[i] ~ dnorm((b1.slope.temp * (x.T[i] - mu.T) + b1.slope.depth * (log(x.depth[i]) - mu.log.depth) + b0.slope)*log(x.ln[i]) + b0.int.depth * (log(x.depth[i]) - mu.log.depth) + b0.int + relake[s.k[i]], sd = sigma)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
p=length(unique(bnla_final$specific_lake_bin)); p
mu.T = mean(bnla_final$avg_temp)
mu.log.depth = mean(log(bnla_final$INDEX_SITE_DEPTH))
mu.log.tp = mean(log(bnla_final$tp))
mu.log.tp.2 = mean(log(bnla_final$tp)^2)

# Prepare data for Nimble model
constants <- list(n = n, 
                  p = p,
                  mu.T = mu.T,
                  mu.log.depth = mu.log.depth,
                  mu.log.tp = mu.log.tp,
                  mu.log.tp.2 = mu.log.tp.2,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  x.T = bnla_final$avg_temp,
                  x.eutro = bnla_final$log_eutro,
                  x.depth = bnla_final$INDEX_SITE_DEPTH,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl)
data <- list(y = yt)
inits <- list(b0.int = 1, b0.int.depth = 1,  b1.R.temp = 1, b1.R.depth = 1, b1.R.tp = 1, b0.R = 1, b1.slope.temp = 1, b1.slope.depth = 1, b0.slope = 1, sigma = 1, s.lake = 1, relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 60000 
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 20000) 

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples)  

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)


# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/pmax((samp1["b1.R.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) + samp1["b1.R.tp","Mean"] * (constants$x.eutro) + samp1["b0.R", "Mean"]), 1))  #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = (samp1["b0.int", "Mean"] + samp1["b0.int.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) +  log(x.ln)*(samp1["b1.slope.temp","Mean"] * (constants$x.T - mu.T) + samp1["b1.slope.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) +  samp1["b0.slope","Mean"])) 

# Calculate R2 and rmse
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 
rmse <- sqrt(mean(y.er^2)); rmse
```

## BVS 2
```{r BVS #2}
# Model Formulation
code <- nimbleCode({
  b0.int ~ dnorm(0, sd = 10)
  b0.R ~ dnorm(20, sd = 10)
  b1.R.depth ~ dnorm(0, sd = 10)
  b1.R.tp ~ dnorm(0, sd = 10)
  b0.slope ~ dnorm(0, sd = 10)
  b1.slope.temp ~ dnorm(0, sd = 10)
  b1.slope.depth ~ dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.lake ~ dunif(0,10)

  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i] , x.tn[i]/max((b1.R.depth * (log(x.depth[i]) - mu.log.depth) + b1.R.tp * (x.eutro[i]) + b0.R), 1))  ### Using original scale for slope of temperature (slope coefficient)
    y[i] ~ dnorm((b1.slope.temp * (x.T[i] - mu.T) + b1.slope.depth * (log(x.depth[i]) - mu.log.depth) + b0.slope)*log(x.ln[i]) + b0.int + relake[s.k[i]], sd = sigma)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
p=length(unique(bnla_final$specific_lake_bin)); p
mu.T = mean(bnla_final$avg_temp)
mu.log.depth = mean(log(bnla_final$INDEX_SITE_DEPTH))
mu.log.tp = mean(log(bnla_final$tp))
mu.log.tp.2 = mean(log(bnla_final$tp)^2)

# Prepare data for Nimble model
constants <- list(n = n, 
                  p = p,
                  mu.T = mu.T,
                  mu.log.depth = mu.log.depth,
                  mu.log.tp = mu.log.tp,
                  mu.log.tp.2 = mu.log.tp.2,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  x.T = bnla_final$avg_temp,
                  x.eutro = bnla_final$log_eutro,
                  x.depth = bnla_final$INDEX_SITE_DEPTH,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl)
data <- list(y = yt)
inits <- list(b0.int = 1, b0.int.depth = 1,  b1.R.temp = 1, b1.R.depth = 1, b1.R.tp = 1, b0.R = 1, b1.slope.temp = 1, b1.slope.depth = 1, b0.slope = 1, sigma = 1, s.lake = 1, relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 60000 
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 20000) 

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)


# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/pmax((samp1["b1.R.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) + samp1["b1.R.tp","Mean"] * (constants$x.eutro) + samp1["b0.R", "Mean"]), 1))  #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = (samp1["b0.int", "Mean"] + log(x.ln)*(samp1["b1.slope.temp","Mean"] * (constants$x.T - mu.T) + samp1["b1.slope.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) +  samp1["b0.slope","Mean"])) 

# Calculate R2 and rmse
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 
rmse <- sqrt(mean(y.er^2)); rmse
```

## BVS 3
```{r BVS #3}
# Model Formulation
code <- nimbleCode({
  b0.int ~ dnorm(0, sd = 10)
  b0.R ~ dnorm(20, sd = 10)
  b1.R.depth ~ dnorm(0, sd = 10)
  b1.R.tp ~ dnorm(0, sd = 10)
  b0.slope ~ dnorm(0, sd = 10)
  b1.slope.temp ~ dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.lake ~ dunif(0,10)

  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i] , x.tn[i]/max((b1.R.depth * (log(x.depth[i]) - mu.log.depth) + b1.R.tp * (x.eutro[i]) + b0.R), 1))  ### Using original scale for slope of temperature (slope coefficient)
    y[i] ~ dnorm((b1.slope.temp * (x.T[i] - mu.T) + b0.slope)*log(x.ln[i]) + b0.int + relake[s.k[i]], sd = sigma)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
p=length(unique(bnla_final$specific_lake_bin)); p
mu.T = mean(bnla_final$avg_temp)
mu.log.depth = mean(log(bnla_final$INDEX_SITE_DEPTH))
mu.log.tp = mean(log(bnla_final$tp))
mu.log.tp.2 = mean(log(bnla_final$tp)^2)

# Prepare data for Nimble model
constants <- list(n = n, 
                  p = p,
                  mu.T = mu.T,
                  mu.log.depth = mu.log.depth,
                  mu.log.tp = mu.log.tp,
                  mu.log.tp.2 = mu.log.tp.2,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  x.T = bnla_final$avg_temp,
                  x.eutro = bnla_final$log_eutro,
                  x.depth = bnla_final$INDEX_SITE_DEPTH,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl)
data <- list(y = yt)
inits <- list(b0.int = 1, b0.int.depth = 1,  b1.R.temp = 1, b1.R.depth = 1, b1.R.tp = 1, b0.R = 1, b1.slope.temp = 1, b1.slope.depth = 1, b0.slope = 1, sigma = 1, s.lake = 1, relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 60000 
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 20000) 

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)


# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/pmax((samp1["b1.R.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) + samp1["b1.R.tp","Mean"] * (constants$x.eutro) + samp1["b0.R", "Mean"]), 1))  #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = (samp1["b0.int", "Mean"] + log(x.ln)*(samp1["b1.slope.temp","Mean"] * (constants$x.T - mu.T) + samp1["b0.slope","Mean"])) 

# Calculate R2 and rmse
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2 
rmse <- sqrt(mean(y.er^2)); rmse
```

## BVS 4
```{r BVS #4}
# Model Formulation
code <- nimbleCode({
  b0.int ~ dnorm(0, sd = 10)
  b0.R ~ dnorm(20, sd = 10)
  b1.R.depth ~ dnorm(0, sd = 10)
  b0.slope ~ dnorm(0, sd = 10)
  b1.slope.temp ~ dnorm(0, sd = 10)
  sigma ~ dunif(0, 10)
  s.lake ~ dunif(0,10)

  for(i in 1:n) {
    x.ln[i] <- min(x.tp[i] , x.tn[i]/max((b1.R.depth * (log(x.depth[i]) - mu.log.depth) + b0.R), 1))  ### Using original scale for slope of temperature (slope coefficient)
    y[i] ~ dnorm((b1.slope.temp * (x.T[i] - mu.T) + b0.slope)*log(x.ln[i]) + b0.int + relake[s.k[i]], sd = sigma)
  }
  for (k in 1:p) {
    relake[k] ~ dnorm(0, sd = s.lake)
  }
})

# Dimension Assignment
n=length(bnla_final$chl); n
p=length(unique(bnla_final$specific_lake_bin)); p
mu.T = mean(bnla_final$avg_temp)
mu.log.depth = mean(log(bnla_final$INDEX_SITE_DEPTH))
mu.log.tp = mean(log(bnla_final$tp))
mu.log.tp.2 = mean(log(bnla_final$tp)^2)

# Prepare data for Nimble model
constants <- list(n = n, 
                  p = p,
                  mu.T = mu.T,
                  mu.log.depth = mu.log.depth,
                  mu.log.tp = mu.log.tp,
                  mu.log.tp.2 = mu.log.tp.2,
                  x.tp = bnla_final$tp,
                  x.tn = bnla_final$tn,
                  x.T = bnla_final$avg_temp,
                  x.eutro = bnla_final$log_eutro,
                  x.depth = bnla_final$INDEX_SITE_DEPTH,
                  s.k = bnla_final$specific_lake_bin)
yt = log(bnla_final$chl)
data <- list(y = yt)
inits <- list(b0.int = 1, b0.int.depth = 1,  b1.R.temp = 1, b1.R.depth = 1, b1.R.tp = 1, b0.R = 1, b1.slope.temp = 1, b1.slope.depth = 1, b0.slope = 1, sigma = 1, s.lake = 1, relake=rep(0,p))

# Build/Run Model and Obtain Samples
Nmodel <- nimbleModel(code, constants = constants, data = data, inits = inits) 
NmodelMCMC <- buildMCMC(Nmodel, monitors2 = c("relake"), enableWAIC = TRUE) 
cNmodel <- compileNimble(Nmodel)
cNmodelMCMC <- compileNimble(NmodelMCMC, project=Nmodel)
n.iter = 60000 
samples <- runMCMC(cNmodelMCMC, niter = n.iter, WAIC = TRUE, nchains = 3, samplesAsCodaMCMC = TRUE, nburnin = 20000) 

# Summary of Output
summary(samples$samples) 

# Rhat Calculations
gelman.diag(samples$samples) 

# Make sample statistics a df mean values can be called directly
samp1 <- data.frame(summary(samples$samples)$statistics)


# Calculate Predictions
x.ln = pmin(constants$x.tp,  constants$x.tn/pmax((samp1["b1.R.depth","Mean"] * (log(constants$x.depth) - mu.log.depth) + samp1["b0.R", "Mean"]), 1))  #limiting nutrient (units of tp)
sum(x.ln==constants$x.tp)/length(x.ln) #fraction of time p is limiting
y.hat = (samp1["b0.int", "Mean"] + log(x.ln)*(samp1["b1.slope.temp","Mean"] * (constants$x.T - mu.T) + samp1["b0.slope","Mean"])) 

# Calculate R2 and rmse
y.er=yt-y.hat
R2=1-sum(y.er^2)/sum((yt-mean(yt))^2); R2
rmse <- sqrt(mean(y.er^2)); rmse
```






